---
title: "Blue Nile Diamonds Analysis"
author: "Group 8: Caroline Bush, Humaira Halim, Leah Hogenmiller, Tatev Kyosababyan"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

### STAT6021 - Linear Models for Data Science

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(RColorBrewer)
library(gridExtra)
library(MASS)
library(patchwork)
```

```{r, include=FALSE}
diamonds <- read.csv("diamonds4.csv")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Factoring cut and clarity so that they are in order from highest quality to lowest quality 
diamonds$cut <- factor(diamonds$cut, levels=c('Astor Ideal', 'Ideal', 'Very Good', 'Good'))
diamonds$clarity <- factor(diamonds$clarity, levels=c('FL', 'IF', 'VVS1', 'VVS2', 'VS1', 'VS2', 'SI1', 'SI2'))
```

<br />

### Section 1

    The diamond is a staple gemstone in jewelry and other luxury items
such that there exists a multitude of prospective buyers, each with a
particular budget. In particular, it is a rite of passage within the
wedding scene to pick the perfect diamond for one's wedding band.
Guiding these buyers through the tradeoffs that allow them to purchase a
high-quality diamond while adhering to their budget requires
comprehensive knowledge of how the characteristics of a diamond
influence its price. In particular, this report describes how the
diamond's carat, cut, color, and clarity each affect price based on the
diamonds that are part of Blue Nile's inventory.<br />     A diamond's
carat, also known as its weight, is one consideration for the
prospective buyer and an important indicator of a diamond's price. Large
diamonds are rarer than smaller diamonds such that most diamonds on the
market are less than two carats. Due to the scarcity of high-carat
diamonds, these gemstones become increasingly more expensive with size:
for every 1% increase in diamond carat, the price increases by 2%. Carat
alone does not dictate a diamond's price, however: cut, color, and
clarity also contribute to a diamond's overall quality. Smaller carat
diamonds are more likely to have high-quality cut, color, and clarity
than larger gemstones. Thus, it is imperative to understand how these
characteristics interact with carat to determine a diamond's total
price.<br />     The cut of a diamond refers to how well-proportioned
the dimensions of the gemstone are, as these properties determine how
light moves and generate the characteristic sparkle that makes diamonds
so attractive. Diamonds with higher cut grades appear more brilliant and
are thought to be priced higher than diamonds of a lower cut grade.
Indeed, these diamonds can reach incredibly lofty prices for their
carat. Opting for a lower cut grade may reduce the price of a diamond,
but that largely depends upon the gemstone's other characteristics. When
considering cut alone, there are diamonds of more modest cut grades with
prices that exceed those of higher-quality cuts. Nevertheless,
prospective buyers working within a budget are often encouraged to
sacrifice carat in favor of a higher-quality cut. Whether this strategy
helps these buyers stay within budget, however, demands consideration of
diamond color and clarity.<br />     After cut, a diamond's color is the
next metric of its quality. Colorless diamonds are the most prized and
also the rarest, which makes them more expensive than diamonds with more
color. As color is more noticeable in diamonds above one carat, the
difference in price across color grades becomes much greater at higher
carats. Prospective buyers considering smaller diamonds are therefore
less likely to see dramatic differences in price across color grades.
Instead, the prices of these diamonds are more strongly influenced by
their cut and clarity. <br />     Clarity is the final metric of diamond
quality included in this report, and is scaled according to the number
and severity of the natural imperfections that occur in all gemstones.
Diamonds with none or only microscopic surface flaws are incredibly
rare, and can be among the most expensive diamonds on the market. Many
imperfections are invisible to the naked eye, however, and can be hidden
by a high-quality cut, particularly among smaller diamonds. As a result,
clarity plays only a minor role in determining the price of most
diamonds on the market, regardless of size. <br />     A diamond's
carat, cut, color, and clarity are among the most important factors
influencing price. Carat is the strongest determinant of price, with
higher-carat diamonds becoming increasingly more expensive regardless of
their cut, color, or clarity. Among diamonds of similar carat, however,
cut plays a large role in dictating price. Diamonds of a high cut
quality tend to have higher prices than those of lower quality cut
grades such that opting for a lower-carat diamond in favor of a better
cut may not necessarily reduce the price. Color and clarity are similar
in that colorless and unblemished diamonds are more expensive than those
of lower color and clarity grades for a given carat. However, this
difference is minimal among most diamonds on the market due to their
small size. These relationships between price and the carat, cut, color,
and clarity of a diamond are important to consider when guiding
prospective buyers towards a high-quality diamond within a particular
budget. <br /> <br />

### Section 2

    Blue Nile's inventory features over 1,000 diamonds of varying
quality as determined by the carat, cut, color, and clarity. These four
qualities are among the most important determinants of a diamond's
price. Using data on the Blue Nile inventory, we have investigated how
specifically these metrics of diamond quality affect the price.

#### Carat

    A diamond's carat is a measure of its weight, with 1 carat
corresponding to 200 mg. Blue Nile states that carat has the largest
effect on price. As carat size increases, diamonds become more
expensive. Indeed, the scatterplot of diamond carat versus price shows
that diamond price increases exponentially with carat. This is due, in
part, to the rarity of large diamond crystals: 95% of diamonds in the
Blue Nile inventory are under 2.1 carats (see histogram below). However,
the bigger is not always better: other characteristics such as cut,
color, and clarity also affect a diamond's quality, and therefore its
total price. <br />

```{r echo=FALSE, out.width = "65%", fig.align = 'center', message=FALSE, warning=FALSE}
# Creating a scatter plot of diamond carat vs price
diamonds %>% 
  ggplot(aes(x=carat, y=price/1000)) +
  geom_point() +
  labs(x='Diamond Carat',y='Price ($1K)', title='Scatterplot of Diamond Carat vs Price')
```

```{r echo=FALSE, out.width = "65%", fig.align = 'center', message=FALSE, warning=FALSE}
# Histogram of the number of diamonds per each 0.25 carats 
diamonds %>% 
  ggplot(aes(x=carat)) + 
  geom_histogram(aes(y=..count../sum(..count..)),binwidth = 0.25, fill = "cornflowerblue")+
  labs(x='Diamond Carat',y='Count', title='Frequency of Diamond Carats')
```

<br />     To explore how other metrics of diamond quality are
represented across carat, we plotted how diamonds of a similar carat are
distributed across the different cut, color, and clarity grades. Blue
Nile states that Astor Ideal is the best and most expensive diamond cut
to purchase as it reflects the most light. The Astor Ideals are more
commonly found at a smaller carat weight, with all the Astor Ideal
diamonds in the inventory being less than 2 carats. There is little
change in the distribution of diamonds across color grades with carat,
which may be due to the fact that Blue Nile sells only high color
quality diamonds. Diamond color grades span D through Z, but Blue Nile's
inventory is limited to diamonds of the D-G (the highest quality) and
H-J (very high quality) color grades. Diamond clarity is similar to cut
as the highest clarity diamonds are limited to those of smaller carat
weights. This presents an interesting point of tension, as inclusions
become more visible in larger diamonds.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of the proportion of diamond cut per carat
cut_per_carat <- diamonds %>% 
  ggplot(aes(x=carat,fill=cut)) +
  geom_histogram(position='fill',binwidth=0.25) +
  labs(x='Diamond Carat', y='Proportion', title='Proportion of Diamond Cut per Carat', fill='Diamond Cut')+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        legend.key.size = unit(0.25, 'cm'),
        legend.key.height = unit(0.25, 'cm'), 
        legend.key.width = unit(0.25, 'cm'), 
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of the proportion of diamond color per carat
color_per_carat <- diamonds %>% 
  ggplot(aes(x=carat,fill=color)) +
  geom_histogram(position='fill',binwidth=0.25) +
  labs(x='Diamond Carat', y='Proportion', title='Proportion of Diamond Color per Carat', fill='Diamond Color')+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        legend.key.size = unit(0.25, 'cm'),
        legend.key.height = unit(0.25, 'cm'), 
        legend.key.width = unit(0.25, 'cm'), 
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of the proportion of diamond clarity per carat
clarity_per_cut <- diamonds %>% 
  ggplot(aes(x=carat,fill=clarity)) +
  geom_histogram(position='fill',binwidth=0.25) +
  labs(x='Diamond Carat', y='Proportion', title='Proportion of Diamond Clarity per Carat', fill='Diamond Clarity')+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        legend.key.size = unit(0.25, 'cm'),
        legend.key.height = unit(0.25, 'cm'), 
        legend.key.width = unit(0.25, 'cm'), 
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8))
```

```{r echo=FALSE, out.width = "110%", fig.align = 'center',message=FALSE, warning=FALSE}
# All 3 histograms of proportion of cut, color and clarity per carat
#cut_per_carat + color_per_carat + clarity_per_cut + plot_layout(ncol=2)
#require(gridExtra)
grid.arrange(cut_per_carat, color_per_carat, clarity_per_cut, ncol=2)
```

    Blue Nile recommends selecting a slightly smaller carat in favor of,
spending more on a good cut, and using shape to mask deficiencies in
color and clarity to purchase the best diamond possible for a given
price. Whether these compromises actually reduce the total price of the
diamond, however, remains to be verified.\
<br / > <br / >

#### Cut

    A diamond's cut determines the amount of light that is reflected out
of it, generating the eye-catching brilliance, fire, and sparkle that
contribute to a diamond's overall beauty. Although most of the public
considers carat the primary factor dictating a diamond's price, Blue
Nile asserts that the cut of a diamond can have the strongest influence.
Prospective buyers are also encouraged to prioritize a superior cut over
a higher carat diamond when shopping within a budget. <br />     There
are six grades of diamond cuts, listed below in descending order of
quality as measured by the amount of light that the diamond reflects:
Astor by Blue Nile™ or Ideal (the top \~3% of diamond cut quality), Very
Good (the top \~15% of diamond cut quality), Good (the top \~25% of
diamond cut quality), Fair, and Poor. All Blue Nile diamonds are graded
by the Gemological Institute of America (GIA), and those of the Astor by
Blue Nile™ cut grade are additionally certified for light performance by
GemEx®. The latter two grades (Fair and Poor) are not sold by Blue Nile,
and so are excluded from this report. <br />     The majority of
diamonds in the Blue Nile inventory are Ideal or Very Good cut grades,
representing 60.87% and 31.47% of diamonds, respectively. The highest
quality diamonds, Astor by Blue Nile™, represent only 1.65% of diamonds
in stock. <br />     

```{r A, out.width = "65%", fig.align = 'center', echo = FALSE, message=FALSE, warning=FALSE}
freq_data <- diamonds %>%
  group_by(cut) %>%
  summarise(Count = n()) %>%
  mutate(Prop = Count/nrow(diamonds))

ggplot(freq_data, aes(x = cut, y = Prop)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(x = "Diamond Cut Grade",  y = "Proportion of Inventory",  
       title = "Distribution of Blue Nile diamonds across cut grades")
```

    The price of a Blue Nile diamond varies widely within each cut
grade, such that it is difficult to glean meaningful insights when all
of the data is plotted on the same set of axes (see below): <br />

```{r B, out.width = "65%", fig.align = 'center', echo = FALSE, message=FALSE, warning=FALSE}
ggplot(diamonds, aes(x = cut, y = price/1000)) + 
  geom_violin() + geom_boxplot(width = 0.1, fill = "cornflowerblue") + ylim(0,375) +
  labs(x = "Diamond Cut Grade",  y = "Price ($1k)",  
       title = "Diamond price by cut")
```

    To more effectively visualize the relationship between a diamond's
price and cut, diamonds with prices above \$10.6k (outliers in this data
set by IQR, $n = 143$) will be graphed separately. The broad trends
observed in these graphs are generally consistent between non-outliers
and outliers, however, so they will be discussed as one. For all
figures, diamonds with prices $< \$10,600$ are graphed on the left, and
diamonds with prices $\geq \$10,600$ are graphed on the right.<br />
    A combination boxplot and violin plot were used to gain a general
knowledge of diamond prices within and between each cut grade. As
expected, diamonds of the Astor by Blue Nile™ cut grade have a higher
median price compared to the other cut grades in the Blue Nile
inventory. The prices of these high-cut quality diamonds are also evenly
distributed across its range. Diamonds of the Ideal and Very Good cut
grades, however, are less consistent in their prices. The highest prices
in these intermediate cut grades exceed that of the most expensive Astor
by Blue Nile™ diamond, but most diamonds in these cut grades are priced
at or below the median value. No diamonds of the Good cut grade quite
reach these extreme values, but they also exhibit an appreciable range
of prices. It is interesting to note that the median price increases
with lower-quality cut grades (excluding Astor by Blue Nile™), despite
Blue Nile's claims that cut is a major determinant of price.<br />

```{r C, out.width = "75%", fig.align = 'center', fig.width=12,echo = FALSE, message=FALSE, warning=FALSE}
# 4646 + 1.5(4646 - 723) = 10530.5; rounded up to 10,600
less_106 <- subset(diamonds, price < 10600)
geq_106 <- subset(diamonds, price >= 10600)
box_v_less <- ggplot(less_106, aes(x = cut, y = price/1000)) + 
  geom_violin() + geom_boxplot(width = 0.05, fill = "cornflowerblue") + 
  labs(x = "Diamond Cut Grade",  y = "Price ($1k)",  
       title = "Price by cut")

box_v_more <- ggplot(geq_106, aes(x = cut, y = price/1000)) + 
  geom_violin() + geom_boxplot(width = 0.05, fill = "cornflowerblue") + 
  labs(x = "Diamond Cut Grade",  y = "Price ($1k)",  
       title = "")

grid.arrange(box_v_less, box_v_more, nrow = 1)
```

<br />     Cut grade is not the sole determinant of a diamond's price:
carat is often viewed as the primary factor influencing price. The
scatterplots below show the relationship between price and carat, while
distinguishing between the different cut grades. A diamond's price
increases with carat, and larger carat diamonds are increasingly more
expensive. This trend holds regardless of the diamond's cut, and
complies with the claims made by Blue Nile on their website. For a given
carat, diamonds of a higher grade cut also tend to be more expensive, in
agreement with Blue Nile's assertion that cut is a key determinant of a
diamond's price. This trend appears to be exaggerated for higher carat
diamonds, albeit imperfectly. <br />

```{r E, echo = FALSE, fig.align = 'center',fig.width=9.5, message=FALSE, warning=FALSE}
scat_less <- ggplot(less_106, aes(x = carat, y = price/1000, color = cut)) + 
  geom_point(alpha = 0.6, size = 3) + 
  scale_color_manual(values = c("blue","gold", "violet", "cornflowerblue"), guide = "none") +
  labs(x = "Diamond Carat",  y = "Price ($1k)",  
       title = "Price by carat and cut")

scat_more <- ggplot(geq_106, aes(x = carat, y = price/1000, color = cut)) + 
  geom_point(alpha = 0.6, size = 3) +
  scale_color_manual(name = "Cut Grade", values = c("blue","gold", "violet", "cornflowerblue")) +
  labs(x = "Diamond Carat",  y = "Price ($1k)", title = "") +
  theme(legend.key.size = unit(0.35, 'cm'), #change legend key size
        legend.title = element_text(size=10), #change legend title font size
        legend.text = element_text(size=8))

grid.arrange(scat_less, scat_more, nrow = 1, widths = c(1.5,2))
```

<br />     Deviations from the general trends outlined above may be
explained by other factors contributing to a diamond's overall value,
namely their color and clarity. While these attributes are discussed in
greater depth below, here we examine how these attributes are
distributed within each cut grade.<br />     Diamonds of the
highest-quality color grades (G and above) appear approximately evenly
represented among Ideal and Very Good cut grade diamonds. Astor by Blue
Nile™ diamonds are most frequently in the F color grade, but this skew
may be a by-product of a lower sample size ($n = 20$). Similarly, an
unexpectedly high proportion of Good cut grade diamonds have the highest
color ranking, but this may be due to the lower sample size ($n = 73$).
<br />     The highest clarity diamonds are extremely rare such that
they make up only a small proportion of the diamonds in each cut grade,
including Astor by Blue Nile™. Of the clarity grades that are in the
Blue Nile inventory, the proportional representation of the higher
clarity grades generally declines with the cut grade as the distribution
of clarity grades shifts somewhat towards lower grades. <br />

```{r F, fig.align = 'center',fig.width=9.5, message=FALSE, warning=FALSE, echo = FALSE}
new_data2 <- as.data.frame(prop.table(table(diamonds$color, diamonds$cut), 2))
new_data2 <- rename(new_data2, color = Var1, cut = Var2)
col_gr <- ggplot(new_data2, aes(x = cut, y = Freq, fill = color)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_brewer(palette = "Set1", name = "Color") +
  labs(x = "Diamond Cut",  y = "Proportion",  title = "Diamond Color by Cut") +
  theme(legend.key.size = unit(0.35, 'cm'), #change legend key size
        legend.title = element_text(size=10), #change legend title font size
        legend.text = element_text(size=8),
        axis.text.x = element_text(angle = 45, vjust = 0.75, hjust = 0.75))

new_data1 <- as.data.frame(prop.table(table(diamonds$clarity, diamonds$cut), 2))
new_data1 <- rename(new_data1, clarity = Var1, cut = Var2)
clar <- ggplot(new_data1, aes(x = cut, y = Freq, fill = clarity)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_brewer(palette = "Set1", name = "Clarity") +
  labs(x = "Diamond Cut",  y = "Proportion",  title = "Diamond Clarity by Cut") +
  theme(legend.key.size = unit(0.35, 'cm'), #change legend key size
        legend.title = element_text(size=10), #change legend title font size
        legend.text = element_text(size=8),
        axis.text.x = element_text(angle = 45, vjust = 0.75, hjust = 0.75))

grid.arrange(col_gr, clar, nrow = 1)
```

<br /> <br />

#### Color

<br />     Blue Nile claims the diamond's color is the second most
important factor, with the ideal diamond being nearly colorless. Lesser
color is associated with a higher grade (assigned a letter D-Z). Diamond
prices increase with an alphabetical order. Finding a diamond without
any color (aka a D-F grade) is incredibly rare, presumably making it
pricier. In contrast, near-colorless diamonds receive a grade between
G-H or I-J. Lastly, faintly colored diamonds would receive a K, which
would be the least expensive grade. Blue Nile does not produce diamonds
past a K grade as per their quality standards. <br />     Given that
color is a categorical variable, we will be visualizing the relationship
between price and color using boxplots in order to see the range of
prices by color. Price is also enumerated by the \$1000's for easier
readability.<br />

```{r, echo = FALSE,out.width = "65%", fig.align = 'center', message=FALSE, warning=FALSE}
ggplot(diamonds, aes(x=color, y=price/1000, fill='yellow')) + 
    geom_boxplot() +
    theme(legend.position="none",plot.title = element_text(hjust = 0.5)) +
    labs(x="Color", y="Price (in $1000's)",
    title="Color over Price of Diamond")

```

    The large quantity of outliers makes it difficult to see the
boxplots themselves, but the extreme range of outliers is informative
itself, with the maximum values decreasing steadily going down the
alphabet. This is consistent with Blue Nile's claim that lesser
color+higher grades are associated with higher prices, and lower grades
being cheaper. <br / >     Beyond the simple relationship between color
and price, Blue Nile also claims that color is more visible in larger
diamonds. Therefore, they claim color matters more when going over 1
carat whereas diamonds under 1 carat are more acceptable for an I-K
grade. To take this relationship into account, we can also visualize the
relationship between all three by creating a scatter plot of carat vs.
price and coloring each point by its diamond color. <br />

```{r, echo = FALSE,out.width = "65%", fig.align = 'center', message=FALSE, warning=FALSE}
diamonds %>%
    ggplot(aes(x=carat, y=price/1000, col=color)) + 
    geom_point(aes(size=carat),shape=18,alpha=0.5) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(x="Carat", y="Price (in $1000's)",
    title="Carat + Color over Price of Diamond")
```

    When looking at this scatter plot, the carat and color become more
and more distinguished as the carat increases. The colors themselves
also begin to form a trend, with higher grade colors being pricier than
lower grade colors. At the lower end of carat ($>2$), the color grades
become less distinguishable, which is also in line with Blue Nile's
claim that color is a less important characteristic when going under $1$
carat. <br /> <br />

#### Clarity

<br />     Diamonds are made underground through enormous pressure and
heat, which means that natural surface flaws are unavoidable. Diamond
clarity is the assessment of small imperfections on the surface and
within the stone. The flaws on the surface of a diamond are called
blemishes; these do not affect the appearance of the diamonds to the
naked eye. These flaws are also considered as "internal characteristics"
since they give the diamond its own character and uniqueness. <br />
    Diamond clarity spans 6 categories with a total of 11 clarity
grades. The lowest grades are I1, I2, and I3, which are the Included
Diamonds--these have visible imperfections likely to affect the beauty
of the diamond. Next there are the SI1, and SI2, the Slightly Included
diamonds, where the inclusions are noticeable at 10x magnification only.
The Very Slightly Included diamonds have minor inclusions either
difficult (VS1) or somewhat easy (VS2) to see at 10x magnification.
These are the most popular ones at Blue Nile, according to the website.
VVS1 and VVS2 represent the Very Very Slightly Included category of the
diamond clarity grade, where even the trained eye cannot spot the
imperfections under 10x magnification. IF stands for the Internally
Flawless diamonds which can have some small surface blemishes visible
under a microscope. And the highest grade for diamond clarity is the
Flawless diamonds (F), which has no internal or external
characteristics. The latter is extremely rare, and can occur with only
1% of the diamonds. <br />      The Blue Nile claims that although
clarity impacts the appearance of a diamond, a high clarity grade does
not guarantee more sparkle or avoidance of imperfections. They suggest
selecting a diamond with inclusions that can't be seen through the crown
without magnification as opposed to choosing one with the highest
clarity grade. The majority of diamond clarity levels in the Blue Nile
diamonds dataset are identified as "Very Slightly Included" or "Slightly
Included", making up about 70% of the entire dataset. Furthermore, the
Included clarity level diamonds are not a part of this dataset.<br />

```{r, echo = FALSE,out.width = "65%", fig.align = 'center', message=FALSE, warning=FALSE}
freq_clarity <- diamonds %>%
  group_by(clarity) %>%
  summarise(Count = n()) %>%
  mutate(Proportions = Count/nrow(diamonds))

ggplot(freq_clarity, aes(x = clarity, y = Proportions)) +
  geom_bar(stat = "identity", fill = "skyblue3") +
  labs(x = "Diamond Cut Grade",  y = "Proportion of Inventory",  
       title = "Distribution of Blue Nile diamonds across cut grades")
```

<br />     Generally, diamonds with the fewest and smallest blemishes
are priced higher, as we can see in the graph below.

```{r, echo = FALSE,out.width = "65%", fig.align = 'center', message=FALSE, warning=FALSE}
ggplot(diamonds, aes(x = clarity, y = price/1000)) + 
  geom_boxplot() + geom_boxplot() + 
  labs(x = "Diamond Clarity Level",  y = "Price ($1k)",  
       title = "Price by Clarity")
```

<br /> However, if we take a look at the datapoints of only flawless
diamonds, which is the highest grade for clarity, we will see that the
clarity level does not necessarily impact the price as much as the other
characteristics of a diamond.<br />

```{r, echo=FALSE, message=FALSE, warning=FALSE}
flawless <- subset(diamonds, clarity == "FL")
flawless
```

<br /> <br /> <br />

### Section 3

<br />     A simple linear regression model of diamond price against
carat was fit using the Blue Nile inventory data. Plotting diamond carat
versus price, there looks to be an exponential relationship between the
two variables. This violates the first assumption of simple linear
regression, which is that there is a linear relationship between the two
variables. Furthermore, a regression plot reveals that the variance in
the residuals increases and is not equally distributed around zero,
violating another assumption of simple linear regression. The residuals
of the model appear to be independent of one another according to an ACF
plot, but a QQ plot indicates that they do not follow a normal
distribution.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Linear regression on original data of carat vs price
results <- lm(price~carat, data=diamonds)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Scatter plot of diamond carat vs price with linear regression line
plot_1 <- diamonds %>% 
  ggplot(aes(x=carat, y=price/1000)) +
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x='Diamond Carat',y='Price ($1K)', title='Scatterplot of Diamond Carat vs Price')+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Residual plot for original data 
res_plot_1 <- diamonds %>% 
  ggplot(aes(x=carat, y=price)) +
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="Fitted y", y="Residuals", title="Residual Plot")+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# ACF plot of original data 
acf_1 <- acf(results$residuals, plot=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# QQ plot using ggplot for original data 
qq_1 <- diamonds %>% 
  ggplot(aes(sample=results$residuals)) +
  stat_qq() +
  stat_qq_line(color='red') +
  labs(x='Theoretical Quantiles', y='Sample Quantiles', title='Normal Q-Q Plot') +
  theme(plot.title = element_text(hjust = 0.5, size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) 
```

```{r echo=FALSE, fig.width=9.5,fig.align = 'center',message=FALSE, warning=FALSE}
# Combining scatter plot, residual plot, qq plot and acf plot of original data together 
grid.arrange(plot_1,res_plot_1,qq_1,ncol=3)
#plot_1 + res_plot_1 + qq_1 + plot_layout(ncol=2, heights = c(4,6)) + 
 # ~plot(acf_1, main=NA, xlab=NA, ylab=NA)+title(main = list("ACF Plot", cex = 0.75),xlab = list("LAG", cex = 0.75), ylab = list("ACF", cex = 0.75))

```

<br />     To address the violations of the assumptions for simple
linear regression discussed above, the data will be transformed. The
response variable will be transformed to help stabilize the variance and
may address the non-linearity of the data. . If the resulting model
fails to meet the assumptions of simple linear regression, then the
predictor variable will also be transformed. <br />     A Box-Cox plot
helps to identify the appropriate transformation of the response
variable. Since the 95% interval is centered close to zero, a log
transformation of the response variable, price, is appropriate. This
also preserves the interpretability of the model coefficients. <br />
    

```{r, echo = FALSE,out.width = "65%", fig.align = 'center', message=FALSE, warning=FALSE}
# Box Cox of original data to determine lambda
boxcox(results, lambda = seq(-1,1,0.5))
```

<br />     After the log transformation on the response variable, the
variance among the residuals is more consistent, and the values are more
evenly distributed around zero. However, as both the scatterplot and the
residual plot show a nonlinear relationship, the assumptions of simple
linear regression are still not being met.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Log Transformation of y into ystar and adding to data frame
ystar<-log(diamonds$price)

diamonds<-data.frame(diamonds,ystar)
```

```{r echo=FALSE,  fig.align = 'center', message=FALSE, warning=FALSE}
# Scatter plot of diamond carat vs transformed y (ystar)
plot_2 <- diamonds %>% 
  ggplot(aes(x=carat, y=ystar)) +
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x='Diamond Carat',y='Price (ystar)', title='Diamond Carat vs Price (ystar)')+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Linear regression on ystar and adding residuals to data frame
result.ystar<-lm(ystar~carat, data=diamonds)
yhat2<-result.ystar$fitted.values
res2<-result.ystar$residuals
diamonds<-data.frame(diamonds,yhat2,res2)
```

```{r echo=FALSE, fig.align = 'center', message=FALSE, warning=FALSE}
# Residual plot of ystar data
res_plot_2 <- diamonds %>% 
  ggplot(aes(x=yhat2, y=res2)) +
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="fitted y", y="Residuals", title="Residual Plot with ystar")+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r echo=FALSE, fig.align = 'center', message=FALSE, warning=FALSE}
# Combining scatter plot and residual plot after transformation on y
grid.arrange(plot_2, res_plot_2, ncol = 2)
```

<br />     A log transformation of the predictor variable, carat, will
be performed in order to fix the nonlinear relationship <br />

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Log transformation on x and adding xstar to data frame
xstar<-log(diamonds$carat)

diamonds<-data.frame(diamonds,xstar)
```

```{r echo=FALSE,  fig.align = 'center', message=FALSE, warning=FALSE}
# Scatter plot of xstar vs ystar
plot_3 <- diamonds %>% 
  ggplot(aes(x=xstar, y=ystar)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE)+
  labs(x='Diamond Carat (xstar)',y='Price (ystar)', title='Diamond Carat(xstar) vs Price(ystar)') +
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Linear Regression model on xstar and ystar and then adding residuals to data frame
results.xstar<-lm(ystar~xstar, data=diamonds)

yhat3<-results.xstar$fitted.values
res3<-results.xstar$residuals

diamonds<-data.frame(diamonds,yhat3,res3)
```

```{r echo=FALSE,  fig.align = 'center', message=FALSE, warning=FALSE}
# Residual plot of xstar and ystar data
res_plot_3 <- diamonds %>% 
  ggplot(aes(x=yhat3,y=res3))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="Fitted y", y="Residuals", title="Residual Plot with xstar and ystar")+
  theme(plot.title = element_text(size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r echo=FALSE, fig.align = 'center', out.width = "65%", message=FALSE, warning=FALSE}
# ACF plot of residuals after transformations
#acf(res3)
AutoCorrelation <- acf(res3, plot = FALSE)
plot(AutoCorrelation, main = "ACF Plot of Residuals After Transformations")
```

```{r, echo=FALSE, fig.align = 'center', fig.width=9.5, message=FALSE, warning=FALSE}
grid.arrange(plot_3, res_plot_3,ncol=2)
```

```{r echo=FALSE, fig.align = 'center', out.width = "65%", message=FALSE, warning=FALSE}
# QQ plot using ggplot for xstar and ystar
qq_2 <- diamonds %>% 
  ggplot(aes(sample=results.xstar$residuals)) +
  stat_qq() +
  stat_qq_line(color='red') +
  labs(x='Theoretical Quantiles', y='Sample Quantiles', title='Normal Q-Q Plot with xstar and ystar') +
  theme(plot.title = element_text(hjust = 0.5, size=10),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) 
qq_2
```

<br />     After performing a log transformation on both the response
and predictor variables, all assumptions of simple linear regression
have been met: there is a linear relationship between the response and
predictor, the variance among the residuals is constant and centered
around zero, the residuals are independent and follow a normal
distribution. Thus, we may now use the estimated regression equation to
draw conclusions about the impact of carat weight on diamond price.
<br /> <br />     The estimated linear regression equation is:
$$y = 8.512 + 1.944x.$$ Here, $y$ is the natural log-transformed price
and $x$ is the natural log-transformed carat. Thus for every one
percent, or one point, increase in diamond carat, there is a 1.944%
increase in the price. This agrees with the original scatterplot, which
indicates that bigger diamonds are exponentially more expensive than
smaller diamonds. Based on this analysis, Blue Nile's suggestion to buy
a slightly smaller diamond when operating on a budget is an effective
strategy. According to the $R^2$, this linear model explains 95.47% of
the variance in the price of Blue Nile diamonds, which lends support to
the assertion that carat is the foremost determinant of a diamond's
price.
